\begin{abstract}
This study proposes a temporal modeling framework with a counterfactual policy-simulation layer to examine student dropout in higher education using time-stamped learning-management-system (LMS) engagement data and administrative withdrawal records. Dropout is operationalized as an observed time-to-event outcome at the course-run enrollment level, and weekly risk is modeled in discrete time via a penalized, class-balanced logistic regression over weekly person--period rows.

Under a late-event temporal holdout, the discrete-time model attains row-level AUCs of 0.8350 on training and 0.8405 on test for discriminating weekly hazard, with aggregate calibration acceptable but sparsely supported in the highest-risk bins. Complementary ablation analyses indicate that performance and survival diagnostics are sensitive to the feature set composition, highlighting the role of temporal engagement signals in risk modeling. For structural evaluation, a hazard-scaling policy is examined via counterfactual simulation, producing scenario-indexed survival contrasts $\Delta S(T)$ under explicit parameters. In the reported run, positive contrasts are confined to the shock branch and remain scenario-dependent, while the mechanism-aware branch is weak to negative. Beyond aggregate evaluation, the framework is applied in a subgroup-sensitive analysis using gender to compare how the same scenario changes survival gaps, with uncertainty quantified via bootstrap; the reported subgroup contrast is directionally stable but very small in magnitude. These results are not presented as causally identified effects, but as evidence that the framework supports internal structural scenario comparison when intervention effects are not identifiable from the available observational data.
\end{abstract}