\section{Appendix B: Reproducibility and Code Availability}
\label{app:b}

All experiments reported in this paper were executed through the project notebook
\texttt{TCM-Student-Dropout\_updated.ipynb} and its exported artifact workflow.
The complete source code and the notebook used to reproduce preprocessing,
temporal split construction, model training/calibration,
policy simulation, subgroup diagnostics, and artifact export are available at
\url{https://github.com/rafa-rodriguess/TCM-Student-Dropout}
\cite{RafaRepoTCMStudentDropout}.

The repository includes instructions to regenerate the exported artifacts
from raw OULAD inputs, including the full \texttt{outputs\_v2} package.
This package contains both \texttt{outputs\_v2/figures} and
\texttt{outputs\_v2/tables}, and all figures and tables referenced in the paper
correspond to files produced by this export process.

The pipeline records artifact provenance during export,
along with key protocol decisions such as split settings,
random seeds, and scenario parameters.
In particular, the reported policy and horizon settings are explicitly tracked,
including $r^*=1$, weekly checking frequency, $W=2$,
a scenario-indexed shock catalog ($\delta_{\mathrm{shock}}\in\{0.08,0.20,0.60\}$ with anchored and hypothetical statuses),
the shared mechanism-aware schedule (week-0 uplift $0.35$, week-1 uplift $0.10$),
the reference scenario identifier used for legacy-compatible outputs,
and the dual-horizon evaluation rule
($T_{policy}=18$, $T_{eval\_metrics}=37$, with
$T_{eval\_policy}=38$ as raw support horizon).

The scenario contract and its traceability fields are exported in
\nolinkurl{table_policy_scenarios_main.csv},
\nolinkurl{table_policy_scenario_params.csv},
\nolinkurl{table_policy_deltaS_by_week_by_scenario.csv}, and
\nolinkurl{table_policy_deltaS_at_horizons_by_scenario.csv}.

For the current manuscript, \texttt{outputs\_v2} is the canonical evidence package, and reported numerical results are aligned to that export set.

Reproducibility is defined at the level of the specified pipeline: the same raw inputs, protocol settings, and export workflow produce the evidence package used in this manuscript.

\section*{AI Disclosure}
Generative AI tools were used for language assistance (translation and wording
suggestions) and for brainstorming/editing support.
No generative AI was used to generate the study data, analyses,
results, or conclusions.
All manuscript content was reviewed and approved by the authors.
